Abstractions for Multi-Sorted Substitutions

### 2023-09-15 paper/paper.tex
Update.

% \documentclass[sigplan, screen]{acmart}
% \documentclass[sigplan, screen, anonymous, review, authordraft]{acmart}
% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\documentclass[sigplan,10pt]{acmart}

@@

\begin{document}
  \begin{abstract}
    Variable binding and substitution are a long-standing nuisance for mechanized
    metatheory of programming languages. A common approach is to represent
    variables as DeBruijn Indices, where substitution is defined in terms of
    variable renaming. If a language has $n$ different kinds of variables, and
    $m$ different kinds of terms, we potentially need to define both
    renaming and substitution for all $m\cdot n$ combinations and prove the
    theory of those operations both in isolation and with respect to each other.

    % McBride introduced kits to define both substitution and renaming by a
    % single term traversal for intrinsically typed terms.\cite{mcbride2005kits}

    We propose a technique to derive the implementation and metatheory of
    renaming and substitution for all $m\cdot n$ combinations of variables and terms
    from a single term traversal and three simple lemmas for extrinsically typed
    but intrinsically scoped terms.
    We demonstrate this technique by formalizing type soundness of System F with
    very parallel substitutions, which replace both type and term variables at
    once, enabling a single substitution-preserves-typing lemma.
    We provide a well-documented implementation of the technique as an Agda
    library.

    As the technique is based on extrinsic typing, the metatheory of renaming
    and substitution is decoupled from type preservation.
    This has the downside, that type preservation of renaming and substitution
    are not for free, but also the upside, that the library can be directly
    instantiated for object languages with non-standard typing relations like
    dependent types, type state, or resource environments.
  \end{abstract}

@@

  \section{Introduction}
  Formalizing metatheory of programming languages in a proof assistant 
  requires to deal with variables, binders and substitution.

  In most languages those constructs behave uniformly,
  i.e. substitution is only concerned with variables and binders and
  behaves homomorphic on all other syntax constructors.

  Intrinsic typing

  \begin{enumerate}
  \item DeBruijn
  \item Intrinsic vs Extrinsic
  \item Reduction under binder
  \end{enumerate}

@@

  Contrary to the usual intrinsic typing, we limit the environment
  \ACode{S} to sorts with index \ACode{\ACon{Var}}.
  This ensures that it is impossible to construct kind variables, as
  it is not possible to extend \ACode{S} with the sort of kinds
  \ACode{\ACon{𝕜}}.

@@

  \section{Map Composition}
  In this section we extend our framework with the composition of
  arbitrary maps.

@@

  To fill this gap, we introduce a new abstraction that we call a
  \ACode{\ACon{ComposeKit}},
  which provides an operation \ACode{\AField{\_\&/⋯\_}} that unifies
  \ACode{\ADef{\_\&\_}} and \ACode{\ADef{\_⋯\_}}, allowing us to
  to define a general composition in the spirit of:
  \ExampleCode\KComposition

@@

  \section{Types \& Contexts}
  \subsection{Library}
  \LibCode\KTypeSorts
  \LibCode\KTypes
  \LibCode\KContextHelper
  \LibCode\KContexts
  \LibCode\KContextLookup
  \subsection{System F}
  \AppCode\FTypes

  \section{Typing}
  % TODO move to typing section
  The only drawback we found, is that types can now also be indexed
  with free expression variables, even though types can never contain
  free variables as assured by the variable constructor. However, this
  only causes minor issues.

  \subsection{Library}
  \LibCode\KVariableTyping
  \LibCode\KTyping
  \LibCode\KTypingKit
  \LibCode\KMapTyping
  \LibCode\KLiftTyping
  % \LibCode\KLiftTypingProof
  \LibCode\KSingleTyping
  % \LibCode\KSingleTypingProof
  \LibCode\KTypingNotation
  \LibCode\KTypingTraversal
  \LibCode\KTypingInstances
  % \LibCode\KTypingTraversalNotation
  \subsection{System F}
  \AppCode\FTyping
  \AppCode\FTypingInst
  \AppCode\FPreserve
  \AppCode\FTypingTraversal

  \section{Semantics}
  % \AppCode\FValues
  \AppCode\FReduction

  \section{Subject Reduction}
  \AppCode\FSubjectReduction
  \AppCode\FSubjectReductionProofInteresting
  % \FSubjectReductionProofRest

  \section{Reflection \& Generics}

>>>>

% \documentclass[sigplan, screen]{acmart}
% \documentclass[sigplan, screen, anonymous, review, authordraft]{acmart}
% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\documentclass[sigplan,10pt, anonymous]{acmart}

@@

\begin{document}
  \begin{abstract}
    Formalizing a typed programming language in a proof assistant
    requires to choose representations for variables and typing.
    Variables are often represented as de Bruijn indices, where
    substitution is usually defined in terms of renamings to allow for
    proofs by structural induction.
    Typing can be represented extrinsically by defining untyped terms and a typing
    relation, or intrinsically by combining syntax and typing into a single
    definition of well-typed terms.
    For extrinsic typing, there is again a choice between extrinsic
    scoping, where terms and the notion of free variables are defined
    separately, and intrinsic scoping, where terms are indexed by their
    free variables.

    This paper describes an Agda framework for formalizing programming
    languages with extrinsic typing, intrinsic scoping, and de Bruijn
    Indices for variables.
    The framework supports object languages with arbitrary many
    variable sorts and dependencies, making it suitable for
    polymorphic languages and dependent types.
    Given an Agda definition of syntax and typing, the framework derives
    substitution operations and lemmas for untyped terms, and provides an
    abstraction to prove type preservation of these operations with
    just a single lemma.
    The key insights behind the framework are the use of multi-sorted syntax
    definitions, which enable parallel substitutions that replace
    all variables of all sorts simultaneously, and
    abstractions that unify the definitions, compositions, and typings
    of multi-sorted renamings and substitutions.
    Case studies have been conducted to prove subject reduction for
    System F with subtyping, dependently typed lambda calculus, and lambda
    calculus with pattern matching.
  \end{abstract}

  % \begin{abstract}
  %   Formalizing a typed programming language in a proof assistant
  %   requires to choose between an extrinsic and an intrinsic
  %   representation of typing.
  %   \emph{Extrinsic} typing means to separately define a set of
  %   untyped terms and a typing relation.
  %   \emph{Intrinsic} typing means to directly define a set of well-typed terms,
  %   incorporating the typing relation into the syntax definition.

  %   One consequence of intrinsic typing is that operations on terms
  %   are also proofs that the operations preserve typing.
  %   This often leads to less code, as shared structure between
  %   definitions and preservation lemmas does not need to be repeated,
  %   but has the drawback that specification and proofs are
  %   intertwined, making it more difficult to check if a specification is
  %   correct.
  % \end{abstract}

  % \begin{abstract}
  %   Formalizing a typed programming language in a proof assistant
  %   requires to choose between an extrinsic and intrinsic representation of typing.
  %   For extrinsic typing, 
  %   There are two approaches to formalizing the syntax of typed object languages
  %   in proof assistants. The extrinsic approach defines
  %   object terms and typing relation separately, whereas the
  %   instrinsic approach only defines well-typed terms.

  %   This paper describes an Agda framework for deriving substitution
  %   operations and lemmas for syntaxes with extrinsic typing,
  %   intrinsic scoping, and DeBruijn Indices as variable representation.

  %   This paper describes an Agda framework for formalizing programming
  %   languages with extrinsic typing, intrinsic scoping, and DeBruijn
  %   Indices as variable representation. The key feature of the
  %   framework 
  %   all sorts of object variables.
  %   Agda's reflection mechanism is used to derive 
  %   substitution properties for a
  %   large class of syntaxes and only requires user code for to prove that the

  %   framework is to deal with multiple sorts of variables in a uniform way
  % \end{abstract}

  % \begin{abstract}
  %   Variable binding and substitution are central concepts in the formalization
  %   of programming languages in proof assistants. 
  %   % Many different representations of variables have been developed with different
  %   % trade-offs on proof simplicity, clarity, and reduction of boilerplate.
  %   A common approach is to represent variables as DeBruijn Indices
  %   and to define substitution in terms of variable renamings for the sake
  %   of structural induction.
  % \end{abstract}

  % \begin{abstract}
  %   Variable binding and substitution are a long-standing nuisance for mechanized
  %   metatheory of programming languages. A common approach is to represent
  %   variables as DeBruijn Indices, where substitution is defined in terms of
  %   variable renaming. If a language has $n$ different kinds of variables, and
  %   $m$ different kinds of terms, we potentially need to define both
  %   renaming and substitution for all $m\cdot n$ combinations and prove the
  %   theory of those operations both in isolation and with respect to each other.

  %   % McBride introduced kits to define both substitution and renaming by a
  %   % single term traversal for intrinsically typed terms.\cite{mcbride2005kits}

  %   We propose a technique to derive the implementation and metatheory of
  %   renaming and substitution for all $m\cdot n$ combinations of variables and terms
  %   from a single term traversal and three simple lemmas for extrinsically typed
  %   but intrinsically scoped terms.
  %   We demonstrate this technique by formalizing type soundness of System F with
  %   very parallel substitutions, which replace both type and term variables at
  %   once, enabling a single substitution-preserves-typing lemma.
  %   We provide a well-documented implementation of the technique as an Agda
  %   library.

  %   As the technique is based on extrinsic typing, the metatheory of renaming
  %   and substitution is decoupled from type preservation.
  %   This has the downside, that type preservation of renaming and substitution
  %   are not for free, but also the upside, that the library can be directly
  %   instantiated for object languages with non-standard typing relations like
  %   dependent types, type state, or resource environments.
  % \end{abstract}

@@

  \section{Introduction}
  In recent years intrinsic typing\cite{TODO} has become rather popular.
  A consequence of intrinsic typing is that operations on terms
  are also proofs that the operations preserve typing.
  This often has the benefit of more concise code, as shared structure
  between definitions and preservation lemmas does not need to be
  repeated.
  However, intrinsic typing also has three major drawbacks:
  \begin{enumerate}
  \item
    It is more difficult to judge whether a specification is correct,
    because there is no clear separation between specification and proofs.
    To understand the specification one has to additionally
    understand the proofs, which can introduce significant cognitive overhead.
    % For example specifying a small step semantics is directly intertwined
    % with the proof of subject reduction.
  \item
    It is less uniform for object languages with multiple variable
    sorts, e.g. type- and expression-variables. In such a scenario,
    types are usually indexed by kinds and expressions by kinds and types.
    This requires not only a type-in-type and expression-in-expression
    substitution, but also a type-in-expression substitution, which causes
    interactions between different types of substitutions introducing
    additional complexity.
    The duplication increases even more if one additionally introduces
    kind-variables. Then a total of 6 different substitutions and
    their interactions is required.
    Unifying these substitutions to reduce repetition is difficult
    with intrinsic typing due to the interdependence of definitions.
  \item
    It is less suitable for automation, because type preservation is baked
    into the syntax definition and type preservation can become arbitrary complex
    depending on the typing relation. This can prevent substitition
    definition and lemmas from being derived, where otherwise it would be possible.
  \end{enumerate}
  With extrinsic typing the major drawback is the duplication between
  definitions and type preservation lemmas, which we found more
  suitable to be eliminated by automation.

  To combat those drawbacks, we went in the direction of intrinsically
  scoped syntax in our search for a general, clean, and automated framework
  for metatheory.

@@

  Note that there is no straightforward way to construct a
  multi-sorted syntax with intrinsic typing: in a direct translation,
  the type of terms \ACode{\ACon{\_⊢\_}} would be indexed by itself, which
  most proof assistants forbid to avoid breaking logical consistency.
  We conjecture that an encoding with universe codes is possible, as
  it has been done for intrinsically typed languages with dependent
  types, but those constructions come with their own drawbacks and complexity.

  % Contrary to the usual intrinsic typing, we limit the environment
  % \ACode{S} to sorts with index \ACode{\ACon{Var}}.
  % This ensures that it is impossible to construct kind variables, as
  % it is not possible to extend \ACode{S} with the sort of kinds
  % \ACode{\ACon{𝕜}}.

@@

  \section{Map Composition}

  % - Reason about multiple applications in terms of a single application
  % - e.g. wk-cancels-⦅⦆ or dist-↑-⦅⦆, often important for induction hypothesis
  % - Mcbride uses tactics
  % - We introduce an abstraction similar to kits
  % - ComposeKit has primitives for composition, ComposeTraversal has the property

  %%%%%%%%%%

  In this section, we extend our framework with a single composition
  operator for arbitrary maps.

  Composition \AField{\_·ₖ\_} equips our framework with the power to
  reason about multiple maps \ACode{ϕ₁} and \ACode{ϕ₂} being applied
  to a term in sequence, in terms of a single application:

  The core property of composition is that applying two maps
  \ACode{ϕ₁} and \ACode{ϕ₂} in sequence to a term \ACode{t}, is the same
  as applying their composition \ACode{(ϕ₁ \AField{·ₖ} ϕ₂)} to \ACode{t}, i.e.
  $$
  \ACode{(t \AField{⋯} ϕ₁) \AField{⋯} ϕ₂ \ACon{≡} t \AField{⋯} (ϕ₁ \AField{·ₖ} ϕ₂)}.
  $$

  %%%%%%%%%%

  In this section, we extend our framework with a single composition
  operator for arbitrary maps.

  Composition \AField{\_·ₖ\_} equips our framework with the power to
  reason about multiple maps \ACode{ϕ₁} and \ACode{ϕ₂} being applied
  to a term in sequence:

  \ACode{\AField{⋯-assoc} t ϕ₁ ϕ₂} \ \ : \ \ \ACode{(t \AField{⋯} ϕ₁) \AField{⋯} ϕ₂ \ACon{≡} t \AField{⋯} (ϕ₁ \AField{·ₖ} ϕ₂)}

  For many lemmas, e.g. substitution-preserves-typing in System F,
  it is necessary to apply an equality like \ACode{t ⋯ σ₁ ⋯ σ₂ ≡ t ⋯ σ₂' ⋯ σ₁'}
  where the induction hypothesis requires \ACode{t' ⋯ σ₁'}.

  %%%%%%%%%%

  In this section, we extend our framework with a single composition
  operator for arbitrary maps.

@@

  To fill this gap, we introduce a new abstraction that we call a
  \ACode{\ACon{ComposeKit}},
  which provides an operation \ACode{\AField{\_\&/⋯\_}} that unifies
  \ACode{\ADef{\_\&\_}} and \ACode{\ADef{\_⋯\_}}, allowing us
  to define a general composition in the spirit of:
  \ExampleCode\KComposition

@@

  \section{Types \& Contexts}
  % To define typing contexts and a typing relation, we need to know
  % which sorts of terms play the role of types for other sorts of terms.
  % When modeling a typed language with multi-sorted syntax, both terms
  % and types are represented as terms, but with different sorts.
  % To define typing contexts and a typing relation, we need to record for each term
  % from 
  % describe terms, and ocan be mapped to 
  % In a typed language,

  In the context of multi-sorted syntax, the notation of a type can be
  described as a mapping between sorts.
  For System F, the expression sort \ACode{\ACon{𝕖}} maps to the type sort
  \ACode{\ACon{𝕥}} and the sort \ACode{\ACon{𝕥}} maps to the kind sort
  \ACode{\ACon{𝕜}}.
  To teach our framework about types, we have to instantiate the following structure:
  \LibCode\KTypeSorts
  For System F the instanciation is
  \AppCode\FTypes
  For a nicer interface, we define a type analogously to a term:
  \LibCode\KTypes

  \subsection{Typing Context}
  A typing context is then a function which maps variables to types.
  As we want our framework to support dependent types, the types returned by
  a typing context are allowed to use the domain of the 
  To be able to deal with dependent types, the result types are only allowed
  to refer to the variables that come before 
  \LibCode\KContextHelper
  \LibCode\KContexts
  \LibCode\KContextLookup

  \LibCode\KVariableTyping

  \section{Typing}
  Equipped with a notion of types and type context we are ready to define
  the multi-sorted typing relation for System F, incorporating both
  typing and kinding:
  \AppCode\FTyping
  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing of the lambda abstraction \ACode{\ACon{⊢λ}} weakens the
    type of the codomain \ACode{t₂}.
    This is necessary, because with multi-sorted syntax, types could
    also depend on expressions, so the sub typing derivation accounts for
    another variable which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-scoped due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  and instanciate it for System F:
  \AppCode\FTypingInst

  Next we define \ACode{\ACon{TypingKit}}s, which are analog to
  \ACode{\ACon{Kit}}s, but abstract about variable and term
  \emph{typing}, instead of variables and terms, themselves.

  \LibCode\KTypingKit
  \subsection{Library}
  \LibCode\KMapTyping
  \LibCode\KLiftTyping
  % \LibCode\KLiftTypingProof
  \LibCode\KSingleTyping
  % \LibCode\KSingleTypingProof
  \LibCode\KTypingNotation
  \LibCode\KTypingTraversal
  \LibCode\KTypingInstances
  % \LibCode\KTypingTraversalNotation
  \subsection{System F}
  \AppCode\FPreserve
  \AppCode\FTypingTraversal

  \section{Semantics}
  We define a standard small-step operational semantics for System F
  with full reduction:
  \AppCode\FReduction

  \section{Subject Reduction}
  Proving subject reduction is now straightforward:
  \AppCode\FSubjectReduction
  \AppCode\FSubjectReductionProofInteresting
  We only listed the cases that do not follow directly from the
  induction hypothesis:
  \begin{itemize}
  \item The beta-reduction rule for type-application, \ACode{\ACon{β-Λ}},
    is a direct consequence of substitution preserves typing.
    \ACode{⊢e₂} is the typing for \ACode{e₂}.
    \ACode{⊢⦅ ⊢e₂ ⦆} is the substitution-typing for the singleton substitution 
    \ACode{\AField{⦅ e₂ ⦆}}.
    \ACode{\AField{⊢e₁ ⊢⋯ ⊢⦅ ⊢e₂ ⦆}} states that applying the singleton substitution
    to \ACode{e₁} preserves typing.
  \item The beta-reduction rule for value-application is almost the same,
    but requires us to addionaly apply an equality to the type index.
  \end{itemize}

  % \FSubjectReductionProofRest

  \section{Reflection \& Generics}
  We used Agda's reflection mechanism\cite{TODO} to derive
  instanciations related to all structures for untyped substitition,
  i.e. \ACode{\ACon{Traversal}} and \ACode{\ACon{ComposeTraversal}}.

  To gain insight into the class of object languages supported by our
  reflection algorithm, we modeled a class of object languages using a
  generic programming technique and instanciated the structures
  generically. The class contains all object languages which:
  \begin{enumerate}
  \item Have a variable constructor of type TODO;
  \item Uses subterms only directly (e.g. not in lists);
  \item Subterms only extend the scope-context.
  \end{enumerate}

  Our reflection algorithm derives proofs with exactly the same
  structure as the generic proof, giving high confidence that it covers
  the same class of languages.

  The formal definition of the generic class can be found in
  Appendix~TODO and the actual implementation.

  We were able to describe many languages in this shape, including
  session types and type state.

  We believe restriction 2 is purely technically and can be lifted
  with a more sophisticated reflection algorithm, as parameterized
  datastructures can also be inlined into the syntax definition as terms
  of a new sort.

  We believe restriction 3 can be lifted partially to allow for
  example the encoding of linear types that requires context
  splitting. This has already be done for intrinsical kits\cite{TODO}.
